Sequential algorithms to split and merge ultra-high resolution 3D images

### 2017-08-11 paper.tex
Updates in Sections II and III

Algorithm~\ref{algo:naive-slices} is a particular case of memory
buffering where the amount of available memory equals the maximum size
of a chunk. More buffering can be achieved when the amount of
available memory increases, as shown in
Algorithm~\ref{algo:buffered-slices}.
\begin{algorithm}[h]
  \caption{Buffered merging from slices}
  \label{algo:buffered-slices}
  \begin{algorithmic}[1]
    \STATE sorted\_slices = sort slices by increasing k values
    \STATE initialize buffer
    \FOR{i = 0 ; i \textless n ; i+=1 }
      \STATE slice = sorted\_slices[i]
      \IF{\texttt{sizeof}(buffer)+\texttt{sizeof}(slice) $\geq$ m}
        \STATE write buffer in reconstructed image
        \STATE clear buffer
      \ENDIF
      \STATE read slice and append it to buffer
    \ENDFOR
  \end{algorithmic}
\end{algorithm}
This algorithm writes in the reconstructed image using a single seek
per memory load. Therefore:
\begin{equation}
N_\mathrm{buff\_slices} =  n + \ceil*{\frac{bR}{m}} \label{eq:buff-slices}
\end{equation}
Buffered slices are straightforward to implement, however, their
extension to block merging is not easy. The remainder of this Section
presents our attempts for such a generalization.

@@

Cluster reads are the more direct extension of buffered slices to
blocks: they load multiple blocks in memory, concatenate them in a
buffer and write the buffer in the reconstructed image. Seeking is
reduced compared to naive block merging since contiguous parts of the
buffer will be written without seeking. A given
block is accessed only once.

@@

\begin{figure}
\centering
\def\svgwidth{0.3\columnwidth}
\input{figures/svg/case1-a.pdf_tex}
\def\svgwidth{0.3\columnwidth}
\input{figures/svg/case2-a.pdf_tex}
\def\svgwidth{0.3\columnwidth}
\input{figures/svg/case3.pdf_tex}
\caption{Memory-load configurations in cluster reads, leading to
  different number of seeks. Red blocks need seeking before each of
  their rows ($d^2$ seeks). Blue blocks need seeking before each of
  their slices ($d$ seeks in total). Green blocks need only a single
  seek. Grey, dashed, transparent blocks represent the contiguous
  memory loads and are added for the sake of visualization.}
\label{fig:cluster-reads}
\end{figure}
Our cluster reads algorithm focuses on the three memory load
configurations represented in Figure~\ref{fig:cluster-reads}, that is,
the amount of memory $m'$ used by the algorithm is rounded down to the
closest number of complete blocks (case 1), of complete block rows (case
2) or of complete block slices (case 3). This is in general reasonable
since adding an incomplete row to a set of complete ones multiplies
the number of required seeks by $d$, as illustrated in
Figure~\ref{fig:avoided-configurations}-Left. In some cases though,
rounding $m$ down to $m'$ might increase the number of required
memory loads to a point that the overall number of seeks also
increases. Such cases are, however, slightly unusual and their
complete description requires extensive calculations involving modulo
arithmetic, which we felt were unwieldy to report here.
\begin{figure}
  \centering
\def\svgwidth{0.3\columnwidth}
\input{figures/svg/incomplete-rows.pdf_tex}
\quad \quad \quad
\def\svgwidth{0.3\columnwidth}
\input{figures/svg/overlap.pdf_tex}
\caption{Configurations that increase the number of seeks per
  memory-load and are thus deliberately avoided by cluster
  reads. Left: configuration with incomplete block rows (multiplies
  number of seeks by $d$). Right: configuration with block rows that
  overlap multiple block slices (multiplies number of seeks by $2$).}
\label{fig:avoided-configurations}
\end{figure}

@@

The amount of memory used $m'$ is set as follows in each of the 3 cases:
\begin{equation*}
  m_1' = \frac{Rb}{n}\floor*{\frac{mn}{Rb}};
  m_2' = \frac{Rb}{\sqrt[3]{n}^2}\floor*{\frac{m\sqrt[3]{n}^2}{Rb}};
  m_3' = \frac{Rb}{\sqrt[3]{n}}\floor*{\frac{m\sqrt[3]{n}}{Rb}}
\end{equation*}
The number of seeks performed by cluster reads in each of the three cases is:
\begin{equation*}
  N^i_{\mathrm{CR}} = n + x_ib_i, \quad i \in \llbracket 1, 3\rrbracket,
\end{equation*}
where $x_i$ is the number of memory loads required to reconstruct the
image and $b_i$ is the number of seeks required to write the memory
load. The first $n$ seeks in the equation are required to read all the
blocks once. According to Figure~\ref{fig:cluster-reads}, we have:
\begin{equation*}
  b_1=d^2=\sqrt[3]{\frac{R}{n}}^2 \quad ; \quad b_2=d=\sqrt[3]{\frac{R}{n}} \quad ; \quad b_3=1
\end{equation*}
The
numbers of memory loads required to reconstruct the image are:
\begin{equation*}
  x_1 = \ceil*{\frac{Rb}{\sqrt[3]{n}^2m_1'}}\sqrt[3]{n}^2;
  x_2 = \ceil*{\frac{Rb}{\sqrt[3]{n}m_2'}}\sqrt[3]{n};
  x_3 = \ceil*{\frac{Rb}{m_3'}}
\end{equation*}
Because our algorithm avoids overlapping configurations, $x_1$ is
proportional to the total number of block rows in the image
($\sqrt[3]{n}^2$) and $x_2$ is proportional to the total number of
block slices ($\sqrt[3]{n}$).

Finally, the total number of seeks
performed by cluster reads to reconstruct the image is:
\begin{equation}
N_\mathrm{CR} =
\begin{cases}
  n + \ceil*{\frac{Rb}{\sqrt[3]{n}^2m_1'}}\sqrt[3]{R}^2     & \text{if } m < \frac{Rb}{\sqrt[3]{n}^2}\\[0.4em]
  n + \ceil*{\frac{Rb}{\sqrt[3]{n}m_2'}}\sqrt[3]{R}        & \text{if } \frac{Rb}{\sqrt[3]{n}^2} \leq m < \frac{Rb}{\sqrt[3]{n}}\\[0.4em]
  n + \ceil*{\frac{Rb}{m_3'}}                              & \text{if } \frac{Rb}{\sqrt[3]{n}} \leq Rb < n\\[0.4em]
  n + 1                                                          & \text{otherwise}
\end{cases} \label{eq:seeks-cluster-reads-1}
\end{equation}
It should be noted that $N_{CR}$ is not a continuous function of m,
due to the differences among $b_i$ values.

@@

\begin{algorithm}[h]
  \caption{Buffered merging of blocks with multiple reads}
  \label{algo:multiple-reads}
  \begin{algorithmic}[1]
  \STATE sorted\_blocks = sort blocks by increasing (k,j,i)
  \STATE start\_index = 0 ; end\_index=(m-1)
  \STATE write\_range = (start\_index, end\_index)
  \WHILE{end\_index \textless Rb}
    \STATE initialize buffer
    \FOR{split in sorted\_blocks}
      \IF{split has voxels in write\_range}
        \STATE split\_data = read split
        \STATE in split\_data, extract the rows in write\_range
        \STATE insert rows in buffer
      \ENDIF
    \ENDFOR
    \STATE write buffer to reconstructed\_image
    \STATE start\_index = end\_index + 1 ; end\_index += m
  \ENDWHILE

@@

As illustrated in Figure~\ref{fig:multiple-reads-cases}, we assume
that $m'$ represents an integer number $k$ of sub-rows (case 1,
$k_1<\sqrt[3]{n}$), of complete rows (case 2, $k_2<d$), of tile rows
(case 3, $k_3<\sqrt[3]{n}$), of slices (case 4, $k_4<d$) or of block
slices (case 5, $k_5<\sqrt[3]{n}$). In each of these 5 cases, we
define $v_i$ as follows:
\begin{equation*}
  v_1=db \text{ ; }  v_2=Db \text{ ; } v_3=Ddb \text{ ; } v_4=D^2b \text{ ; } v_5=D^2db
\end{equation*}
so that we have:
\begin{equation*}
k_i=\floor*{\frac{m}{v_i}} \quad \mathrm{and} \quad m_i'=k_iv_i, \quad i \in \llbracket 1,5 \rrbracket
\end{equation*}

The total number of seeks performed by multiple reads in case $i$ is:
\begin{eqnarray*}
  N^i_{\mathrm{MR}} &=& x_i + (x_i-1)b_i +b_i', \quad i \in \llbracket 1,5 \rrbracket\\
  &=& x_i \left(1+b_i\right)-b_i+b_i'
\end{eqnarray*}
where $x_i$ is the total number of memory loads, $b_i$ is the
number of blocks accessed by the first ($x_i$-1) memory loads and
$b_i'$ is the number of blocks access by the last memory load. The
first $x_i$ seeks in the equation are required to write each memory
load (1 seek per memory load). We have:
\begin{equation*}
  x_i = \ceil*{\frac{Rb}{m_i'}}, \quad i \in \llbracket 1,5 \rrbracket
\end{equation*}
and:
\begin{equation*}
b_1=k_1 ; \quad b_2=\sqrt[3]{n} ; \quad b_3 = k_3\sqrt[3]{n} ; \quad b_4=\sqrt[3]{n}^2 ; \quad b_5=k_5\sqrt[3]{n}^2
\end{equation*}
and:
\begin{eqnarray*}
b_1'=\sqrt[3]{n}D^2\text{ mod }k_1; &b_2'=b_2;   & b_3' = \sqrt[3]{n}\left( \sqrt[3]{n}D\text{ mod }k_3\right)\\
                                   &b_4'=b_4;  & b_5'=\sqrt[3]{n}^2\left( \sqrt[3]{n}\text{ mod }k_5\right) 
\end{eqnarray*}
It gives the following expression for $N_\mathrm{MR}$:
\begin{equation*}
  N_\mathrm{MR} =
\begin{cases}
  \ceil*{\frac{Rb}{m_1'}}(k_1+1)-k_1 \\
  + \left( \sqrt[3]{n}D^2 \text{ mod } k_1 \right)& \text{\hspace*{-0.9em}if } d \leq \frac{m}{b} < D\\[0.4em]

@@

Figure~\ref{fig:model-comparison} plots
Equations~\ref{eq:seeks-cluster-reads-1}
and~\ref{eq:seeks-multiple-reads} for different values of $n$. When
Cluster reads are in Case 1 or 2, they may outperform Multiple reads
only for large values of $n$. When Cluster reads are in case 3, they
are equivalent to Multiple reads; indeed, assuming that $Rb$ is an
exact multiple of $m$, Equations~\ref{eq:seeks-cluster-reads-1}
and~\ref{eq:seeks-multiple-reads} both boil down to $n+\frac{Rb}{m}$.

@@

The 5 algorithms presented earlier are implemented in a Python library
called \texttt{sam} (for ``split and merge''). It uses Nibabel~\cite{matthew_brett_2016_60808} for image I/O
and NumPy for array manipulations.

% buffer implementation
The data buffer used in Cluster reads and Multiple reads is
implemented as a Python dictionary where the keys are offsets in the
reconstructed image and the values are NumPy arrays containing the
data starting at this offset. When the memory load is complete,
dictionary entries are written sequentially to the reconstructed
image. In Cluster reads, some seeking might be required between
writes. We implemented a defragmentation procedure for the dictionary
that merges contiguous dictionary entries in a single one, but we
abandoned it as it proved more time-consuming than going through all
the initial entries, due to the overhead of resizing NumPy arrays to
merge entries. In Multiple reads, dictionary entries are always
contiguous in the reconstructed image.  We tried to use a single NumPy
array as a buffer, but we finally abandoned it as inserting data at a
specific position in a NumPy array copies the data in memory, which
increases the runtime and the peak memory consumption.

% Split methods
The implementation of splitting algorithms was greatly facilitated by
the availability of so-called array proxies in Nibabel, which help to
read specific sub-parts of large images. Nibabel's array proxies
essentially provided the buffer implementation for splitting
algorithms. Unfortunately, they are not available to write
data. \todo{Yongping, Valerie: any other comment on split
  implementations?}

% Additional optimizations.
In Multiple reads, for faster reading all the blocks, before the
merging process start, we read all the block headers to get their
position information and calculate each corresponding row index in the
reconstructed image. We only read the block when there is a row in
that writing range (available memory), otherwise, we skip it.
\todo{Valerie, Yongping, any other implementation trick to report?}

>>>>

Algorithm~\ref{algo:naive-slices} is a particular case of memory
buffering where the amount of available memory equals the maximum size
of a chunk. More buffering can be achieved when the amount of
available memory increases, as shown in
Algorithm~\ref{algo:buffered-slices}.
\begin{algorithm}[h]
  \caption{Buffered merging from slices}
  \label{algo:buffered-slices}
  \begin{algorithmic}[1]
    \STATE sorted\_slices = sort slices by increasing k values
    \STATE initialize buffer
    \FOR{i = 0 ; i \textless n ; i+=1 }
      \STATE slice = sorted\_slices[i]
      \IF{sizeof(buffer)+sizeof(slice) $\geq$ m}
        \STATE write buffer in reconstructed image
        \STATE clear buffer
      \ENDIF
      \STATE read slice and append it to buffer
    \ENDFOR
  \end{algorithmic}
\end{algorithm}
This algorithm writes in the reconstructed image using a single seek
per memory load. Therefore:
\begin{equation}
N_\mathrm{buff\_slices} =  n + \ceil*{\frac{bR}{m}} \label{eq:buff-slices}
\end{equation}
Buffered slices are straightforward to implement, however, their
extension to block merging is not easy. The remainder of this Section
presents our attempts for such an extension.

@@

Cluster reads are the more direct extension of buffered slices to
blocks: they load multiple blocks in memory, concatenate them in a
buffer and write the buffer in the reconstructed image. Seeking is
reduced compared to naive block merging since contiguous parts of the
buffer will be written without seeking. A given
block is accessed only once during the whole merging process.

@@

\begin{figure}
\centering
\def\svgwidth{0.3\columnwidth}
\input{figures/svg/case1-a.pdf_tex}
\def\svgwidth{0.3\columnwidth}
\input{figures/svg/case2-a.pdf_tex}
\def\svgwidth{0.3\columnwidth}
\input{figures/svg/case3.pdf_tex}
\caption{Memory-load configurations in cluster reads, leading to
  different number of seeks. Red blocks need seeking before each of
  their rows ($d^2$ seeks). Blue blocks need seeking before each of
  their slices ($d$ seeks). Green blocks need only a single
  seek. Grey, dashed, transparent blocks represent the contiguous
  memory loads and are added for the sake of visualization.}
\label{fig:cluster-reads}
\end{figure}
Our cluster reads algorithm focuses on the three memory load
configurations represented in Figure~\ref{fig:cluster-reads}, that is,
the amount of memory $m'$ used by the algorithm is rounded down to the
closest number of complete blocks (case 1), of complete block rows (case
2) or of complete block slices (case 3). This is in general reasonable
since adding an incomplete row to a set of complete ones multiplies
the number of required seeks by $d$, as illustrated in
Figure~\ref{fig:avoided-configurations}-Left. In some cases though,
rounding $m$ down to $m'$ might increase the number of required
memory loads to a point that the overall number of seeks also
increases. Such cases are, however, slightly unusual and their
complete description requires extensive calculations involving modulo
arithmetic, which we felt were unwieldy to report here.
\begin{figure}
  \centering
\def\svgwidth{0.3\columnwidth}
\input{figures/svg/incomplete-rows.pdf_tex}
\quad \quad \quad
\def\svgwidth{0.3\columnwidth}
\input{figures/svg/overlap.pdf_tex}
\caption{Configurations that increase the number of seeks per
  memory-load and are thus deliberately avoided by cluster
  reads. Left: configuration with incomplete block rows (multiplies
  the number of seeks by $d$ compared to complete block rows). Right:
  configuration with block rows that overlap multiple block slices
  (multiplies the number of seeks by $2$ compared to non-overlapping
  configurations).}
\label{fig:avoided-configurations}
\end{figure}

@@

The amount of memory used $m'$ is set as follows in each of the 3 cases:
\begin{equation*}
  m_1' = \frac{Rb}{n}\floor*{\frac{mn}{Rb}};
  m_2' = \frac{Rb}{\sqrt[3]{n}^2}\floor*{\frac{m\sqrt[3]{n}^2}{Rb}};
  m_3' = \frac{Rb}{\sqrt[3]{n}}\floor*{\frac{m\sqrt[3]{n}}{Rb}}
\end{equation*}
The number of seeks performed by cluster reads in each of the three cases is:
\begin{equation*}
  N^i_{\mathrm{CR}} = n + x_ib_i, \quad i \in \llbracket 1, 3\rrbracket,
\end{equation*}
where $x_i$ is the number of required memory loads and $b_i$ is the
number of seeks required to write a memory load. The first $n$ seeks
in the equation correspond to the reading of all the blocks. According
to Figure~\ref{fig:cluster-reads}, we have:
\begin{equation*}
  b_1=d^2=\sqrt[3]{\frac{R}{n}}^2 \quad ; \quad b_2=d=\sqrt[3]{\frac{R}{n}} \quad ; \quad b_3=1
\end{equation*}
The
numbers of memory loads required to reconstruct the image are:
\begin{equation*}
  x_1 = \ceil*{\frac{Rb}{\sqrt[3]{n}^2m_1'}}\sqrt[3]{n}^2;
  x_2 = \ceil*{\frac{Rb}{\sqrt[3]{n}m_2'}}\sqrt[3]{n};
  x_3 = \ceil*{\frac{Rb}{m_3'}}
\end{equation*}
Because our algorithm avoids overlapping configurations, $x_1$ is
proportional to the total number of block rows in the image
$\left( \sqrt[3]{n}^2 \right)$ and $x_2$ is proportional to the total number of
block slices $\left( \sqrt[3]{n} \right)$.
Finally, the total number of seeks
performed by cluster reads to reconstruct the image is:
\begin{equation}
N_\mathrm{CR} =
\begin{cases}
  n + \ceil*{\frac{Rb}{\sqrt[3]{n}^2m_1'}}\sqrt[3]{R}^2     & \text{if } m < \frac{Rb}{\sqrt[3]{n}^2}\\[0.4em]
  n + \ceil*{\frac{Rb}{\sqrt[3]{n}m_2'}}\sqrt[3]{R}        & \text{if } \frac{Rb}{\sqrt[3]{n}^2} \leq m < \frac{Rb}{\sqrt[3]{n}}\\[0.4em]
  n + \ceil*{\frac{Rb}{m_3'}}                              & \text{if } \frac{Rb}{\sqrt[3]{n}} \leq Rb < n\\[0.4em]
  n + 1                                                          & \text{otherwise}
\end{cases} \label{eq:seeks-cluster-reads-1}
\end{equation}
It should be noted that $N_{CR}$ is not a continuous function of m,
due to the differences among $b_i$ values.

@@

\begin{algorithm}[h]
  \caption{Buffered merging of blocks with multiple reads}
  \label{algo:multiple-reads}
  \begin{algorithmic}[1]
  \STATE sorted\_blocks = sort blocks by increasing (k,j,i)
  \STATE start\_index = 0 ; end\_index=(m-1)
  \STATE write\_range = (start\_index, end\_index)
  \WHILE{end\_index \textless Rb}
    \STATE initialize buffer
    \FOR{block in sorted\_blocks}
      \IF{block has voxels in write\_range}
        \STATE block\_data = read block
        \STATE in block\_data, extract the rows in write\_range
        \STATE insert rows in buffer
      \ENDIF
    \ENDFOR
    \STATE write buffer to reconstructed\_image
    \STATE start\_index = end\_index + 1 ; end\_index += m
  \ENDWHILE

@@

In the complexity analysis, we assume that $m'$ represents an integer
number $k$ of sub-rows (case 1, $k_1<\sqrt[3]{n}$), of complete rows
(case 2, $k_2<d$), of tile rows (case 3, $k_3<\sqrt[3]{n}$), of slices
(case 4, $k_4<d$) or of block slices (case 5, $k_5<\sqrt[3]{n}$), as
illustrated in Figure~\ref{fig:multiple-reads-cases}. In each of these
5 cases, we define $v_i$ as follows:
\begin{equation*}
  v_1=db \text{ ; }  v_2=Db \text{ ; } v_3=Ddb \text{ ; } v_4=D^2b \text{ ; } v_5=D^2db
\end{equation*}
so that we have:
\begin{equation*}
k_i=\floor*{\frac{m}{v_i}} \quad \mathrm{and} \quad m_i'=k_iv_i, \quad i \in \llbracket 1,5 \rrbracket
\end{equation*}

The total number of seeks performed by multiple reads in case $i$ is:
\begin{eqnarray*}
  N^i_{\mathrm{MR}} &=& x_i + (x_i-1)b_i +b_i', \quad i \in \llbracket 1,5 \rrbracket\\
  &=& x_i \left(1+b_i\right)-b_i+b_i'
\end{eqnarray*}
where $x_i$ is the total number of memory loads, $b_i$ is the
number of blocks accessed by the first $(x_i-1)$ memory loads and
$b_i'$ is the number of blocks access by the last memory load. The
first $x_i$ seeks in the equation correspond to the writing of all memory
loads (1 seek per memory load). We have:
\begin{equation*}
  x_i = \ceil*{\frac{Rb}{m_i'}}, \quad i \in \llbracket 1,5 \rrbracket
\end{equation*}
and:
\begin{equation*}
b_1=k_1 ; b_2=\sqrt[3]{n} ; b_3 = k_3\sqrt[3]{n} ; b_4=\sqrt[3]{n}^2 ; b_5=k_5\sqrt[3]{n}^2
\end{equation*}
and:
\begin{eqnarray*}
b_1'=\sqrt[3]{n}D^2\text{ mod }k_1; &b_2'=b_2;   & b_3' = \sqrt[3]{n}\left( \sqrt[3]{n}D\text{ mod }k_3\right)\\
                                   &b_4'=b_4;  & b_5'=\sqrt[3]{n}^2\left( \sqrt[3]{n}\text{ mod }k_5\right) 
\end{eqnarray*}
It gives the following expression for $N_\mathrm{MR}$:
\begin{equation*}
  N_\mathrm{MR} =
\begin{cases}
  \ceil*{\frac{Rb}{m_1'}}(k_1+1)-k_1 \\
  + \left( \sqrt[3]{n}D^2 \text{ mod } k_1 \right)& \text{\hspace*{-0.9em}if } d \leq \frac{m}{b} < D\\[0.4em]

@@

Figure~\ref{fig:model-comparison} plots
Equations~\ref{eq:seeks-cluster-reads-1}
and~\ref{eq:seeks-multiple-reads} for different values of $n$. When
Cluster reads are in Case 1 or 2, they may outperform Multiple reads
only for large values of $n$. When Cluster reads are in case 3, they
are equivalent to Multiple reads: assuming that $Rb$ is an exact
multiple of $m$, Equations~\ref{eq:seeks-cluster-reads-1}
and~\ref{eq:seeks-multiple-reads} both boil down to $n+\frac{Rb}{m}$.
\todo{We could be a bit more rigorous and detail the cases where
  cluster reads of multiple reads will be better.}

@@

We implemented the 5 algorithms presented earlier in a
Python library called \texttt{sam} (for ``split and merge''). It uses
Nibabel~\cite{matthew_brett_2016_60808} for image I/O and NumPy for
array manipulations.

% buffer implementation
The data buffer used in Cluster reads and Multiple reads is
implemented as a Python dictionary where the keys are offsets in the
reconstructed image and the values are NumPy arrays containing the
data starting at this offset. When the memory load is complete,
dictionary entries are written sequentially to the reconstructed
image. In Cluster reads, some seeking might be required between
writes. We implemented a defragmentation procedure for the dictionary
that merges contiguous dictionary entries in a single one, but we
abandoned it as it proved more time-consuming than going through all
the initial entries, due to the overhead of resizing NumPy arrays to
merge entries. In Multiple reads, dictionary entries are always
contiguous in the reconstructed image.  We tried to use a single NumPy
array as a buffer, but we finally abandoned it as inserting data at a
specific position in a NumPy array copies the data in memory, which
increases both the execution time and the peak memory consumption.

% Split methods
The implementation of splitting algorithms was greatly facilitated by
the availability of so-called array proxies in Nibabel, which help
reading specific sub-parts of large images. Nibabel's array proxies
essentially provided the buffer implementation for splitting
algorithms. Unfortunately, they are not available to write
data.

% Additional optimizations.
In Multiple reads, block headers are read in a first pass where row
indices in the reconstructed image are stored in memory. Those indices
are then processed in each memory load, to identify the blocks that
contribute to it.
